A Bidirected String Graph Model for Genome Assembly

### 2012-12-11 presentation/paper.tex
paper: updates

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
%\usepackage{amsthm}

@@

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technologies can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into fragments, sequencing those fragments, then using a program
called a {\em genome assembler} to put together these short sequences to
reconstruct the original, much longer sequence.  This paper provides an
introduction and formalization of the genome assembly problem, then introduces a
model and algorithm for genome assembly that is based on a special type of graph
called a \KeyTerm{bidirected string graph}.  \end{abstract}

@@

Every living organism contains a collection of hereditary material called a
\KeyTerm{genome}.  Typically, each cell of a multicellular organism contains a
complete copy of that organism's genome.  A genome consists of one or more DNA
(\KeyTerm{deoxyribonucleic acid}) molecules in the form of
\KeyTerm{chromosomes}.  DNA is a double-stranded molecule, each strand of which
is a polymer of \KeyTerm{nucleotides}.  A nucleotide, together with the
corresponding nucleotide on the opposite strand, is called a \KeyTerm{base pair}
(abbreviation: \KeyTerm{bp}).  In DNA, there are four possible nucleotides:
adenine, cytosine, guanine, and thymine, which are abbreviated as \Base{A},
\Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs with T on
the other strand, and \Base{C} always pairs with \Base{G} on the other strand.
In addition, each strand has a direction, and the two strands of a DNA molecule
are directed opposite one another.  Furthermore, a DNA molecule may be circular,
as is usually the case for bacterial chromosomes.

Based on these biological facts, a genome is formally defined, for the purposes
of this paper, as a set of possibly circular, dual-stranded strings of the
symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where dual-stranded has the
meaning indicated above: the two strands run in opposite directions, and the
pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold.

@@

\begin{table}[H]
	\begin{center}
		\begin{tabular}{|l|p{4.5cm}|}
			%\hline
			%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
			%\hline
			%Human mitochondrion &  $\sim$16,000 bp \\
			\hline
			{\bf Organism} & {\bf Approximate genome size (in base
			pairs)} \\
			\hline
			{\it Variola} virus (smallpox) &  186,000 \\
			\hline
			{\it E. coli} (a bacterium) & 4,600,000 \\
			\hline
			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 \\
			\hline
			{\it Zea mays} (domesticated corn) & 2,000,000,000 \\
			\hline
			{\it Homo sapiens} (human)       &  3,200,000,000 \\
			\hline
		\end{tabular}
	\end{center}
	\caption{Approximate genome sizes of selected organisms}
	\label{tab:GenomeSizes}
\end{table}

@@

The \KeyTerm{canonical form} of a DNA sequence is the lexicographically lesser
of that sequence and its reverse-complement.  This concept is useful in
algorithms where a DNA sequence and its reverse-complement are considered
equivalent.

@@

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled together when the overlaps
are used as evidence.  To do this, define a graph where each vertex corresponds
to a read and each edge corresponds to an overlap.  Two vertices are connected
by an edge if and only if the corresponding reads share an overlap.  In
addition, each edge is labeled with an appropriate DNA sequence, the details of
which will be explained later.

It needs to be decided whether each edge will be directed, undirected, or
something else.  To decide this, we need to consider exactly what we want each
edge to actually represent.  Consider three reads $f$, $g$, and $h$, that
overlap as follows:

@@

There are 3 overlaps so there will be 3 edges--- one between $f$ and $g$, one
between $g$ and $h$, and one between $f$ and $h$.  As shown, the reads can be
assembled together consistently when used in the order $f,\ g,\ h$; therefore, it
must be possible to walk $f \to g \to h$ in the graph to model this assembly.
In addition, these reads could actually spell out the reverse-complement
sequence instead:

@@

We might consider a directed graph, but this does not work because the graph
does not allow the reverse-complement walks such as $h \to g \to f$:

\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[->,style=very thick] (f) edge (g);
		\draw[->,style=very thick] (g) edge (h);
		\draw[->,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

Similarly, we might consider an undirected graph, but this does not work because
the graph fails to disallow walks such as $g \to h \to f$ that correspond to
assemblies that are not supported by the underlying reads.

\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[-,style=very thick] (f) edge (g);
		\draw[-,style=very thick] (g) edge (h);
		\draw[-,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

@@

The solution to all these problems is to use a new type of graph called a
\KeyTerm{bidirected graph}.
\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[>->,style=very thick] (f) edge (g);
		\draw[>-<,style=very thick] (g) edge (h);
		\draw[>-<,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

@@

There are 3 (or 4, ignoring symmetry) types of \KeyTerm{bidirected edges}.  They
follow directly from the different types of overlaps (see
\Figure{fig:overlap_mapping}).

@@

The \KeyTerm{in-degree} of a vertex $v$ in a bidirected graph is the number of
heads adjacent to $v$ that point inwards, while the \KeyTerm{out-degree} of $v$
is the number of heads adjacent to $v$ that point outwards.  A bidirected graph
$G$ is \KeyTerm{balanced} if for every vertex in $G$ the in-degree is equal to
the out-degree.

@@

A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
same vertex in such a way that it would be a valid to trace the walk again when
the starting vertex is reached at the end of the cycle.

A bidirected graph is \KeyTerm{connected} if and only if  the undirected graph
produced by replacing all bidirected edges with undirected edges is connected as
defined for undirected graphs.  A bidirected graph is \KeyTerm{strongly
connected} if and only if it is connected and for every edge there is a cycle
containing it.\cite{Medvedev2009}

The concept of a walk in a bidirected graph maps quite nicely onto the genome
assembly problem because a walk through the bidirected string graph constructed
from a set of reads and their overlaps models a way in which the reads can be
assembled together consistently.  In such a walk, each read may be used in
forward or reverse-complement orientation, depending on the direction, relative
to the walk, of the adjacent arrow heads when the corresponding vertex is passed
through.

@@

These interpretations can be generalized to the other two types of bidirected
edges.  Furthermore, these interpretations imply that the actual sequence spelled
out by a walk in the bidirected graph may include implied (non-labeled) sequence
at the start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read).

@@

				\draw[->,style=very thick] (fE) edge (gE);
				\draw[->,style=very thick] (gB) edge (fB);
				\draw (0.90, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.10, -1.01) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
			\end{tikzpicture}
		\end{center}
		\caption{The corresponding directed string graph fragment}
		\label{subfig:bidiedge_interp_a}
	\end{subfigure}

@@

\section{Transitive reduction}

@@

\section{Collapsing unbranched paths}

>>>>

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
%\usepackage{amsthm}

@@

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technologies can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into fragments, sequencing these fragments, then using a program
called a {\em genome assembler} to put together these short sequences to
reconstruct the original, much longer sequence.  This paper provides an
introduction and formalization of the genome assembly problem, then introduces a
model and algorithm for genome assembly that is based on a special type of graph
called a \KeyTerm{bidirected string graph}.  \end{abstract}

@@

Every living organism contains a collection of hereditary material called a
\KeyTerm{genome}.  Typically, each cell of a multicellular organism contains a
complete copy of that organism's genome.  A genome consists of one or more DNA
(\KeyTerm{deoxyribonucleic acid}) molecules in the form of
\KeyTerm{chromosomes}.  DNA is a double-stranded molecule, each strand of which
is a polymer of \KeyTerm{nucleotides}.  A nucleotide, together with the
corresponding nucleotide on the opposite strand, is called a \KeyTerm{base pair}
(abbreviation: \KeyTerm{bp}).  In DNA, there are four possible nucleotides:
adenine, cytosine, guanine, and thymine, which are abbreviated as \Base{A},
\Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs with T on
the opposite strand, and \Base{C} always pairs with \Base{G} on the opposite strand.
In addition, each strand has a direction, and the two strands of a DNA molecule
are directed opposite one another.  Furthermore, a DNA molecule may be circular,
as is usually the case for bacterial chromosomes.

Based on these biological facts, a genome is formally defined, for the purposes
of this paper, as a set of possibly circular, dual-stranded strings of the
symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where dual-stranded has the
meaning indicated above (the two strands run in opposite directions, and the
pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold).

@@

\begin{table}[H]
	\begin{center}
		\begin{tabular}{|c|>{\centering\arraybackslash}p{5.2cm}|}
			%\hline
			%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
			%\hline
			%Human mitochondrion &  $\sim$16,000 bp \\
			\hline
			{\bf Organism} & {\bf Approximate genome size} \\
			\hline
			{\it Variola} virus (smallpox) &  186,000 bp\\
			\hline
			{\it E. coli} (a bacterium) & 4,600,000 bp \\
			\hline
			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 bp \\
			\hline
			{\it Zea mays} (domesticated corn) & 2,000,000,000 bp \\
			\hline
			{\it Homo sapiens} (human)       &  3,200,000,000 bp \\
			\hline
		\end{tabular}
	\end{center}
	\caption{Approximate genome sizes of selected organisms}
	\label{tab:GenomeSizes}
\end{table}

@@

The \KeyTerm{canonical form} of a DNA sequence $S$ is the lexicographically
lesser of $S$ and its reverse-complement $S'$.  This concept is useful in
algorithms where a DNA sequence and its reverse-complement are considered
equivalent.

@@

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled together when the overlaps
are used as evidence.  To do this, define a graph where each vertex corresponds
to a read\footnote{Actually, more than one read may map to the same vertex, as
explained in \Subsection{subsec:removing_contained_reads}.} and each edge
corresponds to an overlap.  Two vertices are connected by an edge if and only if
the corresponding reads share an overlap.  In addition, each edge is labeled
with an appropriate DNA sequence, the details of which will be explained later.

It needs to be decided whether each edge will be directed, undirected, or
something else.  To decide this, we need to consider exactly what each edge
needs to represent.  Consider three reads $f$, $g$, and $h$, that overlap as
follows:

@@

There are 3 overlaps, so there will be 3 edges--- one between $f$ and $g$, one
between $g$ and $h$, and one between $f$ and $h$.  As shown, the reads can be
assembled together consistently when used in the order $f,\ g,\ h$; therefore, it
must be possible to walk $f \to g \to h$ in the graph to model this assembly.
In addition, these reads could actually spell out the reverse-complement
sequence instead:

@@

We might consider a directed graph as shown in
\Subfigure{subfig:possible_directed}, but this does not work because the graph
does not allow the reverse-complement walks such as $h \to g \to f$.

\begin{figure}
	\begin{center}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[->,style=very thick] (f) edge (g);
				\draw[->,style=very thick] (g) edge (h);
				\draw[->,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible directed \\representation of the \\string graph.}
			\label{subfig:possible_directed}
		\end{subfigure}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[-,style=very thick] (f) edge (g);
				\draw[-,style=very thick] (g) edge (h);
				\draw[-,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible undirected \\representation of the \\string graph.}
			\label{subfig:possible_undirected}
		\end{subfigure}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[>->,style=very thick] (f) edge (g);
				\draw[>-<,style=very thick] (g) edge (h);
				\draw[>-<,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible bidirected \\representation of the \\string graph.}
			\label{subfig:possible_bidirected}
		\end{subfigure}
		\caption{Three possibilities for constructing the string graph.  The
		bidirected construction works the best in terms of modeling the
		underlying problem.}
	\end{center}
\end{figure}

Similarly, we might consider an undirected graph as shown in
\Subfigure{subfig:possible_undirected}, but this does not work because
the graph fails to disallow walks such as $g \to h \to f$ that correspond to
assemblies that are not supported by the underlying reads.

@@

The solution to all these problems is to use a new type of graph called a
\KeyTerm{bidirected graph}, shown in \Subfigure{subfig:possible_bidirected}.

@@

There are 3 (or 4, ignoring symmetry) types of \KeyTerm{bidirected edges}.  They
follow directly from the different types of overlaps, as shown in
\Figure{fig:overlap_mapping}.

@@

The \KeyTerm{in-degree} of a vertex $v$ in a bidirected graph is the number of
heads adjacent to $v$ that point inwards (towards $v$), while the
\KeyTerm{out-degree} of $v$ is the number of heads adjacent to $v$ that point
outwards (away from $v$).  A bidirected graph $G$ is \KeyTerm{balanced} if and
only if for every vertex in $G$ the in-degree is equal to the out-degree.

@@

A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
same vertex in such a way that it would be a valid to trace the walk again when
the starting vertex is reached at the end of the walk.

A bidirected graph is \KeyTerm{connected} if and only if  the undirected graph
produced by replacing all bidirected edges with undirected edges is connected
(as defined for undirected graphs).  A bidirected graph is \KeyTerm{strongly
connected} if and only if it is connected and for every edge there is a cycle
containing it\cite{Medvedev2009}.

The concept of a walk in a bidirected graph maps quite nicely onto the genome
assembly problem because a walk through the bidirected string graph constructed
from a set of reads and their overlaps models a way in which the reads can be
assembled together consistently\cite{Myers2005}.  In such a walk, each read may
be used in forward or reverse-complement orientation, depending on the
direction, relative to the walk, of the adjacent arrow heads when the
corresponding vertex is passed through.

@@

These interpretations can be generalized to the other two types of bidirected
edges, and this is described in slightly more detail in \cite{Myers2005}.
Furthermore, these interpretations imply that the actual sequence spelled out by
a walk in the bidirected graph may include implied (non-labeled) sequence at the
start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read).

@@

				\draw[->,style=very thick] (fE) edge (gE);
				\draw[->,style=very thick] (gB) edge (fB);
				\draw (0.90, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.10, -1.01) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
			\end{tikzpicture}
		\end{center}
		\caption{The corresponding directed string graph fragment}
		\label{subfig:bidiedge_interp_c}
	\end{subfigure}

@@

\section{Simplifying the graph}

\subsection{Transitive reduction}

@@

\subsection{Collapsing unbranched paths}
